The Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-06-01 source/sec_problem.tex
Update on Overleaf.
edited assumptions and updated sections 4 and 5

We study the federated learning problem
\begin{align}
	\min _{\mathbf{w}}\left\{F(\mathbf{w}) \triangleq \sum_{k=1}^{N} p_{k} F_{k}(\mathbf{w})\right\},
	\label{eq:problem}
\end{align}
where $N$ is the number of local devices (users/nodes/workers) and $p_k$ is the $k$-th device's weight satisfying $p_k \geq 0$ and $\sum_{k=1}^N p_k = 1$. 
In $k$-th local device, there are a finite number of $n_k$ data points:
$\vx_k^1, \vx_k^2, \dots, \vx_k^{n_k}$.  
The local objective $F_k(\cdot)$ is defined as:
% \begin{align}
$F_{k}(\mathbf{w}) \triangleq \frac{1}{n_{k}} \sum_{j=1}^{n_{k}} \ell\left(\mathbf{w} ; \vx_k^j\right)$,
% \label{eq:localloss}
% \end{align}
where $\ell$ denotes a user-specified loss function. In federated learning, each device stores its local data, which gives rise to its own local objective $F_k$. Note that we do not make any assumption on the
data distributions of each local device. The local minimum
$\vw^{*}_k= \argmin_{\vw \in R^d} F_k(\vw)$ can be drifted far from
the global minimum of \eq{\ref{eq:problem}}, which quantifies the 
heterogeneity of the problem.
In this paper, we systematically investigate the convergence results, 
when the local objectives are convex smooth, strongly convex and convex but
non-smooth.

@@

We make the following standard assumptions on the objective function $F_1,\dots, F_N$. Assumption~\ref{ass:boundedvariance} and Assumption~\ref{ass:subgrad2} have been made by many prior works in federated learning~\cite{yu2019parallel,li2019convergence,stich2018local}. Assumption~\ref{ass:lsmooth} and Assumption~\ref{ass:stroncvx} are common assumptions satisfied by a range of popular objective functions, such as $l-2$
norm regularized logisitc regressions, logistic regressions and cross-entropy loss functions.

\begin{assumption}[Bounded local variance]
	Let $\xi_{t}^{k}$ be sampled from the $k$-th device's local data uniformly at random. The variance of stochastic gradients in each device is bounded: $\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)-\nabla F_{k}\left(\mathbf{w}_{t}^{k}\right)\right\|^{2} \leq \sigma_{k}^{2}$,
	for $k=1, \cdots, N$ and any $\mathbf{w}_{t}^{k}$. Let $\sigma^2=\max_k\sigma_{k}^{2}$.
	\label{ass:boundedvariance}
\end{assumption}
\begin{assumption}[Expected square norm of gradient]
	The expected squared norm of stochastic gradients is uniformly bounded. i.e.,
	$\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)\right\|^{2} \leq G^{2}$, for all $k = 1,..., N$ and $t=0, \dots, T-1$.
	\label{ass:subgrad2}
\end{assumption}
\begin{assumption}[L-smooth]
	$F_{1}, \cdots, F_{N}$ are all $L$-smooth: for all  $\mathbf{v}$  and $\mathbf{w}$, $F_{k}(\mathbf{v}) \leq F_{k}(\mathbf{w})+(\mathbf{v}- \\ \mathbf{w})^{T} \nabla F_{k}(\mathbf{w})+\frac{L}{2}\|\mathbf{v}-\mathbf{w}\|_{2}^{2}$.
	\label{ass:lsmooth}
\end{assumption}
\begin{assumption}[Strongly-convex]
	$	F_{1}, \cdots, F_{N} \text { are all } \mu \text { -strongly convex: for all v and } \mathbf{w}, F_{k}(\mathbf{v}) \geq F_{k}(\mathbf{w})+(\mathbf{v}-\mathbf{w})^{T} \nabla F_{k}(\mathbf{w})+\frac{\mu}{2}\|\mathbf{v}-\mathbf{w}\|_{2}^{2}$
	\label{ass:stroncvx}
\end{assumption}
Assumption~\ref{ass:lsmooth} is made to achieve linear speedup 
of convergence rate with respect to the number of workers, though it
is not necessary for FedAvg and its variants to convergence.
Assumption \ref{ass:subgrad2} implies $\left\| \mathbb{E}\vg_{t,k}\right\|^2  \leq \mathbb{E}\|\vg_{t,k}\|^2 \leq G_k^2$, which further implies $\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)-\nabla F_{k}\left(\mathbf{w}_{t}^{k}\right)\right\|^{2} \leq G^2$, the two assumptions are stated separately because it makes explicit the dependence of the optimality gap on different terms.

>>>>

In this paper, we study the following federated learning problem:
\begin{align}
	\min _{\mathbf{w}}\left\{F(\mathbf{w}) \triangleq \sum\nolimits_{k=1}^{N} p_{k} F_{k}(\mathbf{w})\right\},
	\label{eq:problem}
\end{align}
where $N$ is the number of local devices (users/nodes/workers) and $p_k$ is the $k$-th device's weight satisfying $p_k \geq 0$ and $\sum_{k=1}^N p_k = 1$. 
In $k$-th local device, there are a finite number of $n_k$ data points:
$\vx_k^1, \vx_k^2, \dots, \vx_k^{n_k}$.  
The local objective $F_k(\cdot)$ is defined as:
% \begin{align}
$F_{k}(\mathbf{w}) \triangleq \frac{1}{n_{k}} \sum_{j=1}^{n_{k}} \ell\left(\mathbf{w} ; \vx_k^j\right)$,
% \label{eq:localloss}
% \end{align}
where $\ell$ denotes a user-specified loss function. In federated learning, each device stores its local data, which gives rise to its own local objective $F_k$. Note that we do not make any assumption on the
data distributions of each local device. The local minimum
$\vw^{*}_k= \argmin_{\vw \in R^d} F_k(\vw)$ can be drifted far from
the global minimum of \eq{\ref{eq:problem}}, which quantifies the 
heterogeneity of the problem.
In this paper, we systematically investigate the convergence results, 
when the local objectives are convex smooth, strongly convex and convex but
non-smooth.

@@

We make the following standard assumptions on the objective function $F_1,\dots, F_N$. Assumption~\ref{ass:lsmooth} and Assumption~\ref{ass:stroncvx} are common assumptions satisfied by a range of popular objective functions, such as $\ell^{2}$ regularized logistic regression and cross-entropy loss functions.

 %Assumption~\ref{ass:lsmooth} is made to achieve linear speedup of convergence rate with respect to the number of workers, though it is not necessary for FedAvg and its variants to convergence.

\begin{assumption}[L-smooth]
	$F_{1}, \cdots, F_{N}$ are all $L$-smooth: for all  $\mathbf{v}$  and $\mathbf{w}$, $F_{k}(\mathbf{v}) \leq F_{k}(\mathbf{w})+(\mathbf{v}- \\ \mathbf{w})^{T} \nabla F_{k}(\mathbf{w})+\frac{L}{2}\|\mathbf{v}-\mathbf{w}\|_{2}^{2}$.
	\label{ass:lsmooth}
\end{assumption}
\begin{assumption}[Strongly-convex]
	$	F_{1}, \cdots, F_{N} \text { are all } \mu \text { -strongly convex: for all v and } \mathbf{w}, F_{k}(\mathbf{v}) \geq F_{k}(\mathbf{w})+(\mathbf{v}-\mathbf{w})^{T} \nabla F_{k}(\mathbf{w})+\frac{\mu}{2}\|\mathbf{v}-\mathbf{w}\|_{2}^{2}$
	\label{ass:stroncvx}
\end{assumption}
\begin{assumption}[Bounded local variance]
	Let $\xi_{t}^{k}$ be sampled from the $k$-th device's local data uniformly at random. The variance of stochastic gradients in each device is bounded: $\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)-\nabla F_{k}\left(\mathbf{w}_{t}^{k}\right)\right\|^{2} \leq \sigma_{k}^{2}$,
	for $k=1, \cdots, N$ and any $\mathbf{w}_{t}^{k}$. Let $\sigma^2=\sum_{k=1}^{N}p_k\sigma_{k}^{2}$.
	\label{ass:boundedvariance}
\end{assumption}
\begin{assumption}[Bounded local gradient]
	The expected squared norm of stochastic gradients is uniformly bounded. i.e.,
	$\mathbb{E}\left\|\nabla F_{k}\left(\mathbf{w}_{t}^{k}, \xi_{t}^{k}\right)\right\|^{2} \leq G^{2}$, for all $k = 1,..., N$ and $t=0, \dots, T-1$.
	\label{ass:subgrad2}
\end{assumption}

\textbf{Justfication for Assumptions~\ref{ass:boundedvariance} and~\ref{ass:subgrad2}.} These standard assumptions have been made in many previous
works in federated learning ~\cite{yu2019parallel,li2019convergence,stich2018local}. Here we provide one more justification
for their generality: even though they require bounds for all $\mathbf{w}$,
they are not much stronger than requiring bounds at the global optimal
parameter $\mathbf{w}^{\ast}$. As model average parameters become
closer to $\mathbf{w}^{\ast}$, the $L$-smoothness property implies
that $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}_{t}^{k},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}_{t}^{k})\|^{2}$
will become close to $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})\|^{2}$
and $\mathbb{E}\|\nabla F_{k}(\mathbf{w}^{\ast},\xi_{t}^{k})-\nabla F_{k}(\mathbf{w}^{\ast})\|^{2}$.
Therefore, there is no substantial difference between these assumptions
and assuming the bounds at $\mathbf{w}^{\ast}$ only. On the other
hand, some works such as \cite{haddadpour2019convergence} assume
bounds on the ``gradient diversity'', defined as $\frac{\sum_{k}p_{k}\|\nabla F_{k}(\mathbf{w})\|_{2}^{2}}{\|\sum_{k}p_{k}\nabla F_{k}(\mathbf{w})\|_{2}^{2}}$.
Such assumptions are more restrictive than our assumptions, because
they imply $\nabla F_{k}(\mathbf{w}^{\ast})\equiv0$ for all local
objectives. For such ``overparameterized'' settings, we also prove
that the convergence is in fact exponential.
